---
output:
  github_document:
    html_preview: false
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE, results = "asis", comment = ""}
options(crayon.enabled = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
library(ricu)
old.hooks <- fansi::set_knit_hooks(knitr::knit_hooks)
```

# [ricu](https://eth-mds.github.io/ricu/)

<!-- badges: start -->
[![Lifecycle](https://img.shields.io/badge/lifecycle-stable-brightgreen.svg)](https://lifecycle.r-lib.org/articles/stages.html#stable)
[![CRAN](https://www.r-pkg.org/badges/version/ricu)](https://cran.r-project.org/package=ricu)
[![R build status](https://github.com/eth-mds/ricu/workflows/build/badge.svg)](https://github.com/eth-mds/ricu/actions?query=workflow%3Abuild)
[![R check status](https://github.com/eth-mds/ricu/workflows/check/badge.svg)](https://github.com/eth-mds/ricu/actions?query=workflow%3Acheck)
[![pkgdown build status](https://github.com/eth-mds/ricu/workflows/pkgdown/badge.svg)](https://github.com/eth-mds/ricu/actions?query=workflow%3Apkgdown)
[![covr status](https://github.com/eth-mds/ricu/workflows/coverage/badge.svg)](https://github.com/eth-mds/ricu/actions?query=workflow%3Acoverage)
[![Codecov test coverage](https://codecov.io/gh/eth-mds/ricu/branch/main/graph/badge.svg?token=gQisoami9F)](https://codecov.io/gh/eth-mds/ricu)
<!-- badges: end -->

Working with ICU datasets, especially with publicly available ones as provided by [PhysioNet](https://physionet.org) in R is facilitated by `ricu`, which provides data access, a level of abstraction to encode clinical concepts in a data source agnostic way, as well as classes and utilities for working with the arising types of time series datasets.

## Installation

Currently, installation is only possible from github directly, using the `remotes` if installed

```{r gh-rem, eval = FALSE}
remotes::install_github("eth-mds/ricu")
```

or by sourcing the required code for installation from github by running

```{r gh-dir, eval = FALSE}
rem <- source(
  paste0("https://raw.githubusercontent.com/r-lib/remotes/main/",
         "install-github.R")
)
rem$value("eth-mds/ricu")
```

In order to make sure that some useful utility packages are installed as well, consider installing the packages marked as `Suggests` as well by running

```{r gh-all, eval = FALSE}
remotes::install_github("eth-mds/ricu", dependencies = TRUE)
```

instead, or by installing some of the utility packages (relevant for downloading and preprocessing PhysioNet datasets)

```{r deps-util, eval = FALSE}
install.packages("xml2")
```

and demo dataset packages

```{r deps-data, eval = FALSE}
install.packages(c("mimic.demo", "eicu.demo"),
                 repos = "https://eth-mds.github.io/physionet-demo")
```

explicitly.

## Data access

Out of the box (provided the two data packages `mimic.demo` and `eicu.demo` are available), `ricu` provides access to the demo datasets corresponding to the PhysioNet Clinical Databases eICU and MIMIC-III. Tables are available as

```{r mimic-adm}
mimic_demo$admissions
```

and data can be loaded into an R session for example using

```{r mimic-ts}
load_ts("labevents", "mimic_demo", itemid == 50862L,
        cols = c("valuenum", "valueuom"))
```

```{r mimic-ts-print, echo = FALSE}
load_ts("labevents", "mimic_demo", itemid == 50862L,
        cols = c("valuenum", "valueuom"))[]
```

which returns time series data as `ts_tbl` object.

## Acknowledgments

This work was supported by grant #2017-110 of the Strategic Focal Area "Personalized Health and Related Technologies (PHRT)" of the ETH Domain for the SPHN/PHRT Driver Project "Personalized Swiss Sepsis Study".
