---
title:
  formatted: "\\pkg{ricu}: \\proglang{R}'s interface for ICU data"
  plain:     "ricu: R's interface for ICU data"
  short:     "\\pkg{ricu}: R meets ICU data"
author:
  - name: Nicolas Bennett
    affiliation: ETH Z체rich
    address: >
      Seminar for Statistics
      R채mistrasse 101
      CH-8092 Zurich
    email: \email{nicolas.bennett@stat.math.ethz.ch}
  - name: Drago Plecko
    affiliation: ETH Z체rich
    address: >
      Seminar for Statistics
      R채mistrasse 101
      CH-8092 Zurich
    email: \email{drago.plecko@stat.math.ethz.ch}
  - name: Ida-Fong Ukor
    affiliation: East Kent Hospitals
    address: >
      NHS University Foundation Trust
      William Harvey Hospital
      Kennington Road, Willesborough
      Ashford TN24 0LZ
    email: \email{idafong.ukor@nhs.net}
abstract: >
  The abstract of the article.
keywords:
  formatted: [medicine, intensive care, computational bioinformatics]
  plain:     [medicine, intensive care, computational bioinformatics]
preamble: >
  \usepackage{amsmath}
vignette: >
  %\VignetteIndexEntry{Accessing ICU data with R (Bennett & Plecko, JSS 2020)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output: >
  if (packageVersion("rticles") < 0.5 || rmarkdown::pandoc_version() >= 2)
    rticles::jss_article else rmarkdown::html_vignette
documentclass: jss
bibliography: jss.bib
pkgdown:
  as_is: true
  extension: pdf
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(ricu)
library(data.table)
library(forestmodel)
library(survival)
library(ggplot2)
```

# Introduction

Collection of electronic health records has seen a significant rise in the recent years \cite{evans2016electronic}, opening up opportunities and providing the grounds for a large body of data-driven research oriented towards helping clinicians in decision-making and therefore improving patient care and health outcomes \cite{jiang2017artificial}.

One example of a problem that has received much attention from the machine learning community is early prediction of sepsis in ICU \cite{desautels2016prediction, nemati2018interpretable, futoma2017improved, kam2017learning}. Interestingly, there is evidence that a large proportion of the publications are based on the same dataset \cite{fleuren2019machine}, the Medical Information Mart for Intensive Care (MIMIC) \cite{johnson2016mimic}, which shows a systematic lack of external validation. Part of this problem might well be the lack of a computational infrastructure handling multiple datasets. The MIMIC-III dataset consists of 26 different tables containing about 20GB of data. While much work and care has gone into data preprocessing in order to provide a self-contained ready to use data resource with MIMIC-III, seemingly simple tasks such as computing a SOFA score for patients \citep{vincent1996sofa}, remains a non trivial effort^[There is considerable heterogeneity in number of patients satisfying the Sepsis-3 criterion among studies investigating MIMIC-III. Reported Sepsis-3 prevalence ranges from 11.3% \cite{desautels2016prediction}, over 23.9% \cite{nemati2018interpretable} and 25.4% \cite{wang2018}, up to 49.1% \cite{johnson2018}. While some of this variation may be explained by varying patient inclusion criteria, differences in label implementations must also contribute significantly.]. This is only exacerbated when aiming to co-integrate multiple different datasets of this form, spanning hospitals and even countries, in order to capture effects of differing practice and demographics.

The aim of the \pkg{ricu} package is to provide computational infrastructure allowing users to access complex research questions as easily as possible by providing a unified interface to a heterogeneous set of datasets. The package enables users to write dataset-agnostic code which can simplify implementation and shorten the necessary time prototyping code for different datasets. In its current form, the package handles four large-scale, publicly available intensive care databases out of the box: MIMIC-III from the Beth Israel Deaconess Medical Center in Boston, Massachusetts \cite{johnson2016mimic}, the eICU Collaborative Research Database \cite{pollard2018eicu}, containing data collected from 208 hospitals across the United States, HiRID from the Department of Intensive Care Medicine of the Bern University Hospital, Switzerland \cite{faltys2020hirid} and AmsterdamUMCdb from the Amsterdam University Medical Center \cite{aumc}. Furthermore, \pkg{ricu} was designed with extensibility in mind such that adding further public and or private user-provided datasets is possible.

To our knowledge, infrastructure that provides a common interface to multiple such datasets is a novel contribution. While there have been efforts such as \cite{adibuzzaman2016closing} and \cite{wang2020mimic} attempting to abstract away some specifics of a dataset, these have so far exclusively focused on MIMIC-III, the most popular of public ICU datsets and have not been designed with dataset interoperability in mind.

Given the somewhat narrow focus of the targeted datasets, combined with the fact that in some cases data is even extracted from identical systems, it may come as a surprise as to how heterogeneous the resulting datasets are. In MIMIC-III and HiRID, for example, time-stamps are reported as absolute times (albeit randomly shifted due to data privacy concerns), whereas eICU and AUMC use relative times (with origins being admission times). Furthermore, while relative times in eICU are reported in minutes, AUMC uses millisecond resolution. Another example, involves different types of patient identifiers and their use among datasets. Common to all is the notion on an ICU admission ID, but apart from that, the amount of available information varies. While ICU readmissions for a given patient can be identified in both eICU and MIMIC-III, this is not possible on either HiRID or AUMC. Similarly, individual patients can be identified on AUMC and MIMIC-III, but not on eICU or HiRID. In this sense, MIMIC-III is the most complete dataset of the four, but this comes with issues of its own: ID availability among tables varies, requiring extra processing steps for retrieving data with common IDs.

The following sections serve to introduce \pkg{ricu} by first giving an overview of particularities of the four default dataset, followed by some implementation details on how datasets are stored and how external data sources can be made accessible by \pkg{ricu} (\ref{data_srcs}). The next section is concerned with ...

# Data sources {#data_srcs}

## Ready to use datasets

## Implementation details

## Adding external datasets

# Data concepts

In this Section we go over the categories of data useful for research problems related to intensive care medicine. The categories we define are fairly broad and somewhat loosely defined, as this is not the main focus of the manuscript.

## Physiological data
Labs, vitals. Could introduce the \code{ts_tbl} here.

## Treatment-related information
Antibiotics, vasopressors, mechanical ventilation... could introduce the \code{win_tbl} here.

## Co-morbidities

Based on ICD-9 codes. Should enable the extraction of co-morbidities used for the Charlson and Elixhauser scores.

## Admission diagnoses
Categorizing into surgical, non-surgical and other might be sufficient for now.

## Patient information
Age, gender, other demograpichs, patient stay information.

## Outcomes
Death outcome, prolonged ICU stay outcome.

## Implementation details

# Examples

We focus on two simple examples with which we try to cover most of the data types described in Section \ref{implementation}.

```{r tcm_loader, include = FALSE}
load_concepts <- cached_concept_loader("tcm")
```

## Lactate and mortality
The first example we look at is the association of lactate levels and mortality. This problem has been studied before and it is widely accepted that both static and dynamic lactate indices are associated with increased mortality \citep{haas2016severe, nichol2011dynamic, van2013cumulative}. We quickly look at how one might fit a time-varying proportional hazards Cox model \citep{therneau2015package} in order to investigate this problem. We additionally include the Sequential Organ Failure Assessment (SOFA) score \citep{vincent1996sofa} as a general predictor of illness severity.

```{r time_cox_model}
last_event <- function(x) c(rep(0, length(x)-1), max(x))
# data loading
tbl <- load_concepts(c("lact", "death", "sofa"), "mimic_demo",
                     verbose = FALSE)
tbl <- tbl[, c(meta_vars(tbl), "lact", "sofa", "death"), with = FALSE]
tbl <- tbl[, lact := nafill(lact, "locf")]
tbl <- tbl[, lact := nafill(lact, fill = 1)]
tbl[, event := as.integer(sum(death, na.rm = TRUE) > 0), by = c(id_vars(tbl))]
tbl[, event := last_event(event), by = c(id_vars(tbl))]
tbl[, next_charttime := charttime+1L]
# model fitting
cox_time_mod <- coxph(
  Surv(charttime, next_charttime, event) ~ lact + sofa,
  data = tbl
)
```

We visualize the results of the model

```{r cox_plot, echo = FALSE, warning = FALSE, message = FALSE}
forest_model(cox_time_mod)
```

A simple exploration already shows that the increased values of lactate are associated with mortality, even after adjusting for the SOFA score.

```{r diab_loader, include = FALSE}
load_concepts <- cached_concept_loader("diab")
```

## Diabetes and insulin treatment
The next example we turn to covers the usage of co-morbidities and treatment related information. We look at the amount of insulin administered to patients in the first 24 hours from their ICU admission. In particular, we investigate if patients who are diabetic receive more insulin in the first day of their stay. For this we create two concepts: `ins24`, a binned variable representing the cumulative amount of insulin administered within the first 24 hours of an ICU admission, and `diab`, a logical variable encoding diabetes co-morbidity.

As there already is a concept available, capable of retrieving raw insulin administration, `ins24` can be implemented as `rec_cncpt`, making sure that `ins` is loaded with aggregation set to `sum()` and injecting the callback function `ins_cb()` into the loading process. The callback function takes care of the pre-processing steps outlined above: first data is subsetted to fall into the the first 24 hours of ICU admissions, followed by binning of summed values.

```{r ins24}
ins_breaks <- c(0.01, 10, 20, 30, 40)

ins_cb <- function(ins, ...) {

  day_one <- function(x) x >= hours(0L) & x <= hours(24L)

  idx_var <- index_var(ins)
  ids_var <- id_vars(ins)

  ins <- ins[day_one(get(idx_var)), list(ins24 = sum(ins)),
             by = c(ids_var)]
  ins <- ins[, ins24 := list(
             .bincode(ins24, breaks = c(-Inf, ins_breaks, Inf)) - 1)]

  ins
}

ins24 <- load_dictionary("mimic_demo", "ins")
ins24 <- concept("ins24", ins24, "insulin in first 24h", aggregate = "sum",
                 callback = ins_cb, target = "id_tbl", class = "rec_cncpt")
```

The binary diabetes concept can for example be implemented as `lgl_cncpt`, for which ICD-9 codes are matched using a regular expression. As we're not only interested in retrieving diabetic patients, a `col_itm` is more suited for data retrieval over an `rgx_itm` and for creating the required callback function that produces a logical vector we can use `transform_fun()` coupled with a function like `grep_diab()`. The two concepts are then combined using `c()` and loaded via `load_concepts()`.

```{r diab}
grep_diab <- function(x) grepl("^250\\.?[0-9]{2}$", x)

diab  <- item("mimic_demo", table = "diagnoses_icd",
              callback = transform_fun(grep_diab), class = "col_itm")
diab  <- concept("diab", diab, "diabetes", target = "id_tbl",
                 class = "lgl_cncpt")

dat <- load_concepts(c(ins24, diab), id_type = "icustay", verbose = FALSE)
dat <- replace_na(dat, 0, cols = "ins24")
dat
```

After this, we can visualize the difference between the two groups with a histogram:

```{r diabetes_visualize, echo = FALSE}

bin_labels <- function(breaks, unit, lower0 = TRUE) {

  x_labels <- sapply(1:(length(breaks)-1),
    function(x) paste0("[", breaks[x], "-", breaks[x+1], "]")
  )

  first_label <- paste0("< ", breaks[1])
  if (lower0) first_label <- paste0("[0-", breaks[1], "]")
  x_labels <- c(first_label, x_labels, paste0("> ", breaks[length(breaks)]))

  paste(x_labels, unit)
}

ggplot(dat, aes(x = ins24, fill = diab, colour = diab)) +
  geom_histogram(aes(y = ..count.. / sum(..count..)), alpha = 0.75,
                 position = "dodge", bins = 1 + length(ins_breaks)) +
  xlab("Amount of administered insulin in first 24h of ICU stay") +
  ylab("Proportion of patients") +
  scale_x_continuous(labels = bin_labels(ins_breaks, "units"),
                     breaks = c(0, seq_along(ins_breaks))) +
  theme_minimal(10) +
  theme(legend.position = c(0.7, 0.7),
        legend.background = element_blank(),
        legend.box.background = element_rect(colour = "black"),
        legend.text = element_text(size = 12),
        legend.title = element_text(size = 16))
```

The plot might suggest that diabetic patients do receive more insulin that non-diabetic patients, in the first day of ICU stay.
