---
title: "Quick start guide"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quick start guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(ricu)
```

```{r, echo = FALSE}
demo  <- c("mimic_demo", "eicu_demo")
alls  <- demo

avail <- is_data_avail(alls)

srcs_avail <- function(x) all(avail[x])

paste1 <- function(x) {

  x <- paste0("`", x, "`")

  if (length(x) <= 2L) {
    return(paste0(x, collapse = " and "))
  }

  paste0(paste0(head(x, n = length(x) - 2L), collapse = ", "), ", ",
         paste0(tail(x, n = 2L), collapse = " and "))
}

if (!srcs_avail(alls)) {

  msg <- paste(
    "Note: Examples in this vignette require that datasets",
    paste1(alls[!avail]), "are available. Chunks that depend on certain",
    "datasets will not be evaluated if the corresponding dataset is missing.",
    "The full version of this vignette is available from",
    "https://CRAN.R-project.org/package=ricu/vignettes/ricu.html"
  )

  message(paste(strwrap(msg), collapse = "\n"))
}
```

In order to set up `ricu`, download of datasets from several platforms is required. Two data sources, `mimic_demo` and `eicu_demo` are available directly as R packages, hosted on Github. The respective full-featured versions `mimic` and `eicu`, as well as the `hirid` dataset are available from [PhysioNet](https://physionet.org), while access to the remaining standard dataset `aumc` is available from yet another [website](https://amsterdammedicaldatascience.nl/). The following steps guide through package installation, data source set up and conclude with some example data queries.

## Package installation

Stable package releases are available from [CRAN](https://cran.r-project.org/package=ricu) as

```{r, eval = FALSE}
install.packages("ricu")
```

and the latest development version is available from [GitHub](https://github.com/eth-mds/ricu) as

```{r, eval = FALSE}
remotes::install_github("eth-mds/ricu")
```

## Demo datasets

The two demo datasets `mimic_demo` and `eicu_demo` are listed as `Suggests` dependencies and therefore their availability is determined by the value passed as `dependencies` to the above package installation function. The following call explicitly installs the demo dataset packages

```{r, eval = FALSE}
install.packages(
  c("mimic.demo", "eicu.demo"),
  repos = "https://eth-mds.github.io/physionet-demo"
)
```

## Full datasets

Included with `ricu` are functions for download and setup of the following datasets: `mimic`, `eicu`, `hirid` and `aumc`, which can be invoked in several different ways.

- To begin with, a directory is needed where the data can permanently be  stored. The default location is platform dependent and can be overridden using the environment variable `RICU_DATA_PATH`. The current value can be retrieved by calling `data_dir()`.
- Access to both the PhysioNet datasets (MIMIC-III, eICU and HiRID), as well as to AUMCdb is free but credentialed. In addition to setting up an [account with PhysioNet](https://physionet.org/register/), [credentialing](https://physionet.org/settings/credentialing/) is required, which, in the case of HiRID, must also be followed by submitting an [access request](https://physionet.org/request-access/hirid/1.1.1/) to the data-owners. Details on the procedure for requesting access to AUMCdb is available from [here](https://amsterdammedicaldatascience.nl/#amsterdamumcdb) and consists of filling out a [form](https://amsterdammedicaldatascience.nl/arfeula_v1.6.pdf) and completing a training course such as [Data or Specimens Only Research (DSOR)](https://physionet.org/about/citi-course/) which, together with proof of training course completion, can be submitted by [email](mailto:access@amsterdammedicaldatascience.nl).
- If raw data in `.csv` form has already been downloaded, this can be decompressed and copied to an appropriate sub-folder (`mimic`, `eicu`, `hirid` or `aumc`) to the directory identified by `data_dir()`.
- In order to have `ricu` download the required data, login credentials can be supplied as environment variables `RICU_PHYSIONET_USER`/`RICU_PHYSIONET_PASS` and `RICU_AUMC_TOKEN` (the string the follows `token=` in the download URL received from the AUMCdb data owners) or entered into the terminal manually in interactive sessions.
- Enabling efficient random row/column access, `ricu` converts `.csv` files into a binary format using the [fst](https://cran.r-project.org/package=fst) package.
- Data conversion to `.fst` format (and potentially data download) is automatically triggered upon first access of a table. In interactive sessions, the user is asked for permission to setup the given data source and in non-interactive sessions, access to missing data throws an error.
- Instead of relying on first data access to trigger setup, up-front data conversion, possible preceded by data download, can be invoked by calling `setup_src_data()`.

## Concept loading

Many commonly used clinical data concepts are available for all data sources, where the required data exists. An overview of available concepts is available by calling `explain_dictionary()` and concepts can be loaded using `load_concepts()`:

```{r, eval = srcs_avail(demo)}
head(explain_dictionary(src = c("mimic_demo", "eicu_demo")))
load_concepts("alb", "mimic_demo", verbose = FALSE)
```

Concepts representing time-dependent measurements are loaded as `ts_tbl` objects, whereas static information is retrieved as `id_tbl` object. Both classes inherit from `data.table` (and therefore also from `data.frame`) and can be coerced to any of the base classes using `as.data.table()` and `as.data.frame()`, respectively. Using `data.table` 'by-reference' operations, this is available as zero-copy operation by passing `by_ref = TRUE`

```{r, eval = srcs_avail("eicu_demo")}
(dat <- load_concepts("height", "eicu_demo", verbose = FALSE))
head(as.data.frame(dat, by_ref = TRUE))
```

Many functions exported by `ricu` use `id_tbl` and `ts_tbl` objects in order to enable more concise semantics. Merging an `id_tbl` with a `ts_tbl`, for example, will automatically use the columns identified by `id_vars()` of both tables, as `by.x`/`by.y` arguments, while for two `ts_tbl` object, respective columns reported by `id_vars()` and `index_var()` will be used to merge on.

## Extending the concept dictionary

In addition to the ~100 concepts that are available by default, adding user-defined concepts is possible either as  R objects or more robustly, as JSON configuration files.

- Data concepts consist of zero, one, or several data items per data source, encoding how to retrieve the corresponding data. The constructors `concept()` and `item()` can be used to instantiate concepts as R objects.

    ```{r, eval = srcs_avail("mimic_demo")}
    ldh <- concept("ldh",
      item("mimic_demo", "labevents", "itemid", 50954),
      description = "Lactate dehydrogenase",
      unit = "IU/L"
    )
    load_concepts(ldh, verbose = FALSE)
    ```

- Configuration files are looked for in both the package installation directory and in user-specified locations, either using the environment variable `RICU_CONFIG_PATH` or by passing paths as function arguments (`load_dictionary()` for example accepts a `cfg_dirs` argument).

- Mechanisms for both extending and replacing existing concept dictionaries are supported by `ricu`. The file name of the default concept dictionary is called `concept-dict.json` and any file with the same name in user-specified locations will be used as extensions. In order to forgo the internal dictionary, a different file name can be chosen, which then has to be passed as function argument (`load_dictionary()` for example has a `name` argument which defaults to `concept-dict`)

- A JSON-based concept akin to the one above can be specified as

    ```
    {
        "ldh": {
          "unit": "IU/L",
          "description": "Lactate dehydrogenase",
          "sources": {
            "mimic_demo": [
              {
                "ids": 50954,
                "table": "labevents",
                "sub_var": "itemid"
              }
            ]
          }
        }
    }
    ```

    and this can (given that it is saved as `concept-dict.json` in a directory pointed to by `RICU_CONFIG_PATH`) then be loaded using `load_concepts()` as

    ```{r, eval = FALSE}
    load_concepts("ldh", "mimic_demo")
    ```

    For further details on constructing concepts, refer to documentation at `?concept` and `?item`.
