---
title: "Units of measurement"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Units of measurement}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(ricu)
library(ggplot2)

srcs  <- c("mimic_demo", "eicu_demo")
avail <- is_data_avail(srcs)

srcs_avail <- function(x) all(avail[x])

if (!srcs_avail(srcs)) {

  msg <- paste(
    "Note: Examples in this vignette require that one or more of datasets",
    paste("`", srcs, "`", collapse = ", "), "are available. Chunks that",
    "depend on certain datasets will not be evaluated if the corresponding",
    "dataset is missing. In order to download data, have a look at",
    "`?setup_src_data`."
  )
  msg <- paste(strwrap(msg), collapse="\n")
  message(msg)
}

```

### Possible mismatch in units

Working with different ICU datasets can be challenging in terms of units of measurement. In particular, combining data from different countries can cause a mismatch in the units, as the practices vary substantially. In particular, we note that the commonly used unit of measurement for laboratory values in the US datasets is mg/dL, as opposed to mmol/L used in European datasets. Note that the conversion between the two requires the molecular weight of the substance and therefore must be handled on case-to-case basis. When loading data, care needs to be taken in light of this possible problem.

### `ricu` approach

All concepts that can be loaded with `load_concepts()` within `ricu` have been checked for units, and the units were converted where necessary.

For example, we take 5 different concepts:

```{r concept-range, eval = srcs_avail(srcs)}
srcs

concepts <- c(
  "map", "lact" , "crea", "bili", "plt"
)

dat <- lapply(srcs,
  function(src) load_concepts(concepts, src, verbose = FALSE)
)

names(dat) <- srcs

dat
```

We plot the density of the features and report their median values:

```{r uom, echo = FALSE, fig.width = 6, eval = srcs_avail(srcs)}
take_quants <- function(x, lwr, upr) {
  qq <- quantile(x, probs = c(lwr, upr), na.rm = TRUE)
  x[!is.na(x) & x >= qq[1L] & x <= qq[2L]]
}

for (conc in concepts) {

  feat <- lapply(dat, `[[`, conc)
  meds <- vapply(feat, median, numeric(1L), na.rm = TRUE)
  feat <- Map(data.frame, src = names(feat),
              dat = lapply(feat, take_quants, lwr = 0.05, upr = 0.95))

  title <- paste0(conc, ": median values ",
    paste0(round(meds, 2), " (", names(meds), ")", collapse = ", ")
  )

  print(
    ggplot(Reduce(rbind, feat), aes(x = dat, fill = src)) +
      geom_density(alpha=0.5) +
      xlab(conc) + theme_bw() + ggtitle(title)
  )
}
```

Note that the matching between datasets is not perfect, but the median values should align closely (the above are done using the demo datasets which are rather small in size; in general the matching should be even better).

### Concepts outside `ricu` dictionary

Not all relevant concepts are included in the `ricu` dictionary. When loading concepts outside the dictionary, we recommend checking whether the units match across datasets using the density plots and median values as shown above. In particular, if there is a clear difference in the median values, or if the density plots look "multimodal", there is reason to believe some unit conversion is required.
