---
title: "Units of measurement"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Units of measurement}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(ricu)
library(ggplot2)
```

```{r, echo = FALSE}
demo  <- c("mimic_demo", "eicu_demo")
full  <- c("mimic", "eicu", "aumc", "hirid")

avail <- is_data_avail(c(demo, full))

srcs_avail <- function(x) all(avail[x])

if (!srcs_avail(demo)) {

  msg <- paste(
    "Note: Examples in this vignette require that one or more of datasets",
    paste("`", demo, "`", collapse = ", "), "are available. Chunks that",
    "depend on certain datasets will not be evaluated if the corresponding",
    "dataset is missing. In order to download data, have a look at",
    "`?setup_src_data`."
  )
  msg <- paste(strwrap(msg), collapse="\n")
  message(msg)
}
```

One challenge when working with electronic healthcare (EHR) data from different hospitals, which is accentuated if the intensive care units (ICUs) collecting the data are located in different geographic regions, is the use of different measurement unit systems. In part, this can be attributed to the use of imperial units in English speaking countries (e.g. patient weight being reported in lbs instead of kg), but more subtle differences in practice are involved as well, such as reporting lab test results in mg/dL instead of the SI unit mmol/L. While discrepancies of the former type are easy to resolve, harmonization of different notions of concentration is slightly more involved due to the conversion factor being substance-specific.

## `ricu` data concepts

Data concepts of type `num_cncpt` can specify an expected units as character vector (where the string in position 1 is added as per-column attribute `units`). Due to the complexity involved, unit conversion to the specified target unit currently is not handled automatically^[Support for automatically handling this, using the [units](https://cran.r-project.org/package=units) packages, is being considered, but so far has not been implemented]. Instead, item-specific callback functions are responsible for handling this, while unit mismatches are reported as messages during concept loading.

```{r, eval = srcs_avail(demo)}
(dat <- load_concepts(c("lact", "map"), c("mimic_demo", "eicu_demo")))
```

Messages raised during loading of lactate values, for example, indicate that 0.62% of retrieved values are specified in mEq/L instead of mmol/L (which requires an identity transformation for unit conversion), while the remaining discrepancies are false positives (both mmole/L and MMOLL can be assumed to mean mmol/L). For mean arterial bloop pressure values, the target unit is specified as mmHg (with the alternative spelling mm Hg being accepted as well), however, due to the data organization in eICU^[Both the `vitalperiodic` and `vitalaperiodic` tables in eICU are layed out in *wide* format (whereas most other tables are in *long* format), and therefore no unit columns are available. This also explains the substantial degree of missingness (in terms of values) reported, as such a *wide* data organization scheme coupled with differing measurement intervals per variable will inevitably lead to some degree of sparsity.], no explicit measurement units are specified for this variable, in turn causing the large percentage of missing unit values reported.

Several utility functions are exported from `ricu` for helping with creating callback functions that handle unit conversion. Data items corresponding to the bilirubin concept for the European datasets HiRID and AUMCdb, for example, have a callback entry specified as `convert_unit(binary_op(``*``, 0.058467), "mg/dL")`. This creates a callback functions which applies `binary_op(``*``, 0.058467)` to the column specified as `val_var` and replaces existing values in the column identified by `unit_var` with the value `"mg/dl"`. In case the loaded data already is comprised of a mix of units, a regular expression passed as `rgx` can be specified, which will be used to identify the rows on which to operate. Finally, the function `binary_op` turns a binary function into an unary function by fixing the second operand.

As a sanity check to see whether measurement units (and data values in general) are aligned when compared across datasets, density plots may be employed.

```{r, echo = FALSE, fig.width = 6, eval = sum(avail) >= 2L}
filter_quants <- function(x, lwr, upr) {

  do_filter <- function(x) {
    qq <- quantile(x, probs = c(lwr, upr), na.rm = TRUE)
    x[!is.na(x) & x >= qq[1L] & x <= qq[2L]]
  }

  x[, list(val = do_filter(data_col(x))), by = "source"]
}

srcs <- c(full, demo)
srcs <- head(srcs[is_data_avail(srcs)], n = length(full))

concepts <- c("map", "lact" , "crea", "bili", "plt")

dat <- load_concepts(concepts, srcs, verbose = FALSE, merge = FALSE)
dat <- lapply(dat, filter_quants, lwr = 0.05, upr = 0.95)

for (feat in names(dat)) {

  title <- dat[[feat]][, list(val = median(val)), by = "source"]
  title <- paste0(title$val, " (", title$source, ")", collapse = ", ")
  title <- paste0(feat, ": ", title)

  print(
    ggplot(dat[[feat]], aes(x = val, fill = source)) +
      geom_density(alpha = 0.5) +
      xlab(feat) + theme_bw() + ggtitle(title)
  )
}
```

When extending the `ricu` dictionary to both new data sources and new data concepts, it might be worthwhile to visually inspect the returned data in a fashion similar to the above in order to have a high-level confirmation that measurement units roughly agree.
